---
title: "V-inflated Poisson count regression"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Motivation

We want to model the distribution of groups size in parrot populations.
These animals tend to form smaller or larger groups, but
groups of size 2 are also often observed as a result of mating pairs.

The count distribution is characterized with a spike at 2, and by the
absence of 0s due to group size being conditional on having >0 birds to consider
it a group.

We start developing a general V-Inflated Poisson (VIP) model, then we add the >0 condition.
Simulations are done to check the estimating procedure.

# Maximum likelihood

Let $Y$ be a random variable, and $y$ are observations, $V$ is the count value
that has some extra probability mass ($V=0$ is the ZIP model), $f(y; \lambda)$
is the Poisson density ($f(y; \lambda) = e^{-\lambda} \frac{\lambda^{y}}{y!}$).

The V-Inflated density can be written as $P(Y=y) = \phi I(Y=V) + (1-\phi) f(y; \lambda)$
which is $\phi + (1-\phi) f(V; \lambda)$ when $Y=V$ and
$(1-\phi) f(y; \lambda)$ otherwise.

## Functions

The `vip` function does the optimization. `method` argument can take values
listed for `optim` and also `"DE"` for differential evolution algorithm.
If there are convergence issues with `"Nelder-Mead"`, try `"SANN"` and `"DE"`.

```{r}
library(DEoptim)
library(Matrix)
vip <-
function(Y, X, Z, V=0,
offsetx, offsetz, weights, linkz="logit",
truncate=FALSE, hessian=TRUE, method="Nelder-Mead", init=NULL, ...) {
    if (missing(Y))
        stop("C'mon, you must have some data?!")
    if (truncate && any(Y < 1))
        stop("Y must be >0 when truncate=TRUE")
    n <- length(Y)
    id0 <- Y == V
    id1 <- !id0
    if (missing(X)) {
        X <- matrix(1, n, 1)
        colnames(X) <- "(Intercept)"
    }
    if (missing(Z)) {
        Z <- matrix(1, n, 1)
        colnames(Z) <- "(Intercept)"
    }
    kx <- ncol(X)
    kz <- ncol(Z)
    if (missing(offsetx))
        offsetx <- 0
    if (missing(offsetz))
        offsetz <- 0
    if (missing(weights))
        weights <- rep(1, n)
    linkinvx <- poisson("log")$linkinv
    linkinvz <- binomial(linkz)$linkinv
    good.num.limit <- c(.Machine$double.xmin, .Machine$double.xmax)^(1/3)

    ## VIP model full likelihood
    nll_VIP_ML <- function(parms) {
        mu <- as.vector(linkinvx(X %*% parms[1:kx] + offsetx))
        phi <- as.vector(linkinvz(Z %*% parms[(kx + 1):(kx + kz)] + offsetz))
        loglik0 <- log(phi + (1 - phi) * dpois(V, lambda = mu, log = FALSE))
        loglik1 <- log(1 - phi) + dpois(Y, lambda = mu, log = TRUE)
        loglik <- sum(weights[id0] * loglik0[id0]) + sum(weights[id1] * loglik1[id1])
        if (!is.finite(loglik) || is.na(loglik))
            loglik <- -good.num.limit[2]
        -loglik
    }
    ## 0-truncated VIP model full likelihood
    nll_VIP_TR <- function(parms) {
        mu <- as.vector(linkinvx(X %*% parms[1:kx] + offsetx))
        phi <- as.vector(linkinvz(Z %*% parms[(kx + 1):(kx + kz)] + offsetz))
        loglik0 <- log(phi + (1 - phi) * dpois(V, lambda = mu, log = FALSE) / (1-exp(-mu)))
        loglik1 <- log((1 - phi) * dpois(Y, lambda = mu, log = FALSE) / (1-exp(-mu)))
        loglik <- sum(weights[id0] * loglik0[id0]) + sum(weights[id1] * loglik1[id1])
        if (!is.finite(loglik) || is.na(loglik))
            loglik <- -good.num.limit[2]
        -loglik
    }
    .solvenear <-
    function(x)
    {
        xinv <- try(solve(x), silent = TRUE)
        if (inherits(xinv, "try-error"))
            xinv <- as.matrix(solve(Matrix::nearPD(x)$mat))
        xinv
    }
    if (is.null(init))
        init <- rep(0, kx+kz)
    nll <- if (truncate) nll_VIP_TR else nll_VIP_ML

    if (method == "DE") {
        DElimit <- 10
        up <- rep(DElimit, length(init))
        lo <- -up
        opt <- DEoptim(fn=nll, lower=lo, upper=up,
            control=list(trace=FALSE, itermax=length(init)*200))
        par <- opt$optim$bestmem
        names(par) <- c(paste0("P_", colnames(X)), paste0("V_", colnames(Z)))
        ll <- -opt$optim$bestval
        if (hessian) {
            hess <- optimHess(opt$optim$bestmem, nll)
            vc <- .solvenear(hess)
        } else {
            vc <- matrix(NA, length(par), length(par))
        }
    } else {
        opt <- optim(init, nll,
            hessian=hessian, method=method, ...)
        par <- opt$par
        vc <- if (hessian)
            .solvenear(opt$hessian) else matrix(NA, length(par), length(par))
        ll <- -opt$value
    }
    names(par) <- c(paste0("P_", colnames(X)), paste0("V_", colnames(Z)))
    dimnames(vc) <- list(names(par), names(par))
    out <- list(call=match.call(),
        coefficients=par, 
        loglik=ll, 
        vcov=vc, 
        nobs=n,
        truncate=truncate, 
        Y=Y, 
        X=X, 
        Z=Z, 
        V=V,
        offsetx=offsetx, 
        offsetz=offsetz, 
        weights=weights, 
        linkz=linkz, 
        method=method, 
        init=init)
    class(out) <- "vip"
    out
}
vcov.vip <- function(object, ...) object$vcov
logLik.vip <- function (object, ...)
    structure(object$loglik, df = object$nobs - length(object$coef),
        nobs = object$nobs, class = "logLik")
summary.vip <- function (object, ...) {
    k <- length(object$coefficients)
    coefs <- coef(object)
    se <- sqrt(diag(vcov(object)))
    tstat <- coefs/se
    pval <- 2 * pnorm(-abs(tstat))
    coefs <- cbind(coefs, se, tstat, pval)
    colnames(coefs) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
    coefs <- coefs[1:k, , drop = FALSE]
    rownames(coefs) <- names(coef(object))
    out <- list(call = object$call, coefficients=coefs, loglik = object$loglik,
        bic=BIC(object), truncate=object$truncate)
    class(out) <- "summary.vip"
    return(out)
}
print.summary.vip <- function (x, digits, ...)
{
    if (missing(digits))
        digits <- max(3, getOption("digits") - 3)
    cat("\nCall:", deparse(x$call,
        width.cutoff = floor(getOption("width") * 0.85)), "", sep = "\n")
    cat("V-Inflated", if (x$truncate) "(Zero-Truncated)" else "", "Poisson Model\n\n")
    cat(paste("Coefficients:\n", sep = ""))
    printCoefmat(x$coefficients, digits = digits, signif.legend = FALSE)
    if (!any(is.na(array(x$coefficients)))) {
        if (getOption("show.signif.stars") & any(x$coefficients[,4] < 0.1))
            cat("---\nSignif. codes: ", "0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1", "\n")
    }
    cat("\nLog-likelihood:", formatC(x$loglik, digits = digits),
        "\nBIC =", formatC(x$bic, digits = digits), "\n")
    cat("\n")
    invisible(x)
}
confint.vip <-
function (object, parm, level = 0.95, ...)
{
    cf <- coef(object)
    pnames <- names(cf)
    if (missing(parm)) {
        parm <- pnames
    } else {
        if (is.numeric(parm))
            parm <- pnames[parm]
    }
    a <- (1 - level)/2
    a <- c(a, 1 - a)
    pct <- paste(format(100 * a, trim = TRUE, scientific = FALSE, digits = 3), "%", sep="")
    ci <- array(NA, dim = c(length(parm), 2), dimnames = list(parm, pct))
    fac <- qnorm(a)
    ses <- sqrt(diag(vcov(object, model, type)))
    ci[] <- cf[parm] + ses[parm] %o% fac
    ci
}
```

## Simple case

```{r}
set.seed(123)
n <- 1000
lam <- 2 # poisson mean, can be a vector of length n
phi <- 0.4 # V-inflation probability, can be a vector of length n
V <- 2 # V is the count value, can be 0, 2, etc
y <- y0 <- rpois(n, lam)
a <- rbinom(n, 1, phi)
y[a > 0] <- V
table(Poisson=y0, Vinflated=y)

mod <- vip(Y=y, V=2)
summary(mod)
cbind(True=c(log_lam=log(lam), logit_phi=qlogis(phi)),
      Est=coef(mod))
```

## Covariates for the non-V part

```{r}
set.seed(123)
n <- 1000
x <- rnorm(n)
df <- data.frame(x=x)
X <- model.matrix(~x, df)
beta <- c(-0.5,-0.5) # Intercept and beta values for covariate
lam <- exp(X %*% beta) # poisson mean, can be a vector of length n
phi <- 0.4 # V-inflation probability, can be a vector of length n
V <- 2 # V is the count value, can be 0, 2, etc
y <- y0 <- rpois(n, lam)
a <- rbinom(n, 1, phi)
y[a > 0] <- V
table(Poisson=y0, Vinflated=y)
mod <- vip(Y=y, X=X, V=2)
summary(mod)
cbind(True=c(beta=beta, logit_phi=qlogis(phi)),
      Est=coef(mod))
```

## Methods

```{r}
coef(mod)
vcov(mod)
summary(mod)
confint(mod)
nobs(mod)
logLik(mod)
AIC(mod)
BIC(mod)
```


# Zero-truncated VIP

We can truncate counts to be larger than 0. We also need $V>0$
(for $V=0$ case, look into ZIP or conditional Poisson model).
Conceptually, the V-Inflation follows the
0-truncation (because we cannot observe 0, real truncated distribution).

The 0-truncated PDF is $P(Y=y \mid Y>0) = \frac{P(Y=y)}{1 - P(Y=0)}$.
The 0-truncated V-Inflated density is
$P(Y=y  \mid Y>0,V>0) = \phi I(Y=V) + (1-\phi) \frac{f(y; \lambda)}{1-f(0; \lambda)}$.
This can be achieved in the `vip` call by the argument `truncate=TRUE`.

Here we use covariates for both the V and non-V part.

```{r}
set.seed(1)
n <- 1000
x <- rnorm(n)
z <- runif(n, -1, 1)
df <- data.frame(x=x, z=z)
X <- model.matrix(~x, df)
Z <- model.matrix(~z, df)
beta <- c(-0.5, -0.5)
alpha <- c(0, 0.5)
lam <- exp(X %*% beta)
phi <- plogis(Z %*% alpha)
V <- 2 # V is the count value, cannot be 0
y <- y0 <- rpois(n, lam)
a <- rbinom(n, 1, phi)
keep <- y0>0
y <- y[keep] # conditioning (i.e. exclude 0s)
y0 <- y0[keep]
X <- X[keep,]
Z <- Z[keep,]
y[a[keep] > 0] <- V
table(Poisson=y0, Vinflated=y)

mod <- vip(Y=y, X=X, Z=Z, V=2, truncate=TRUE)
summary(mod)
cbind(True=c(beta=beta, alpha=alpha),
      Est=coef(mod))
```

# Goodness of fit

```{r}
goodness <- function(object, maxcount=NULL) {
    if (is.null(maxcount))
        maxcount <- max(object$Y)
    COUNTS <- if (object$truncate)
        1L:max(object$Y) else 0L:maxcount
    
    P_obs <- table(object$Y) / nobs(object)
    P_obs <- as.numeric(P_obs[match(COUNTS, names(P_obs))])
    P_obs[is.na(P_obs)] <- 0
    names(P_obs) <- COUNTS
    
    P_exp <- P_obs
    P_exp[] <- 0
    
    linkinvx <- poisson("log")$linkinv
    linkinvz <- binomial(object$linkz)$linkinv
    parms <- coef(object)
    kx <- ncol(object$X)
    kz <- ncol(object$Z)
    mu <- as.vector(linkinvx(object$X %*% parms[1:kx] + object$offsetx))
    phi <- as.vector(linkinvz(object$Z %*% parms[(kx + 1):(kx + kz)] + object$offsetz))
    #id0 <- object$Y == object$V
    Pmat <- matrix(0, nobs(object), length(COUNTS))
    colnames(Pmat) <- COUNTS
    
    PV <- if (object$truncate) {
        log(phi + (1 - phi) * dpois(object$V, 
            lambda = mu, log = FALSE) / (1-exp(-mu)))
    } else {
        log(phi + (1 - phi) * dpois(object$V, lambda = mu, log = FALSE))
    }
    for (i in COUNTS) {
        PC <- if (object$truncate) {
            log((1 - phi) * dpois(i, lambda = mu, log = FALSE) / (1-exp(-mu)))
        } else {
            log(1 - phi) + dpois(i, lambda = mu, log = TRUE)
        }
        Pmat[,as.character(i)] <- if (i == object$V)
            exp(PV) else exp(PC)
    }
    P_exp <- colMeans(Pmat)
    cbind(P_obs=P_obs, P_exp=P_exp)
}

## fit null model
mod0 <- vip(Y=y, X=X[,1,drop=FALSE], Z=Z[,1,drop=FALSE], V=2, truncate=TRUE)
## calculate GoF for null and other model
(gof0 <- goodness(mod0))
(gof <- goodness(mod))
## better model is closer to the 1:1 line
plot(gof0, type="b", col=2)
points(gof, type="b", col=4)
abline(0, 1, lty=2)
## better model will give smaller absolute deviation
sum(abs(apply(gof0, 1, diff)))
sum(abs(apply(gof, 1, diff)))
```

